Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-08-04 paper-correlations.tex
half of discussion i hope

\bibliographystyle{elsarticle-num}
\journal{Physica A: Statistical Mechanics and its Applications}

@@

\author[Smith]{Vasily~Postnicov}
\author[Smith,MIT]{Aleksei~Samarin}
\author[Smith]{Marina~Karsanina}
\author[Smith]{Aleksey~Khlyupin}
\author[Smith]{Kirill~Gerke}
\ead{kg@ifz.ru}

\address[Smith]{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 107031, Russia}
\address[MIT]{Computational Mathematics and Cybernetics, Lomonosov Moscow State
  University, Moscow, 119991, Russia}

@@

There are different types of correlation functions that in general describe some
probabilities. The main parameter of any correlation function is the number of
points that are utilized to evaluate such a probability. While so-called
$n$-point probability function \cite{Torquato_book} will completely describe any
structure in the limit of $n \rightarrow$ number of voxels/pixels on the image,
computing or storing such a function is not practical. Based on information
content, it was shown that increasing the number of point $n > 2$ only
marginally improves the quality of the structure quantification and these
improvements decay with increasing $n$ \cite{yao1993high,Gommes2}. For stochastic reconstruction purposes
computation of higher-order statistics will, arguably, not be balanced by
increased number of points, but this is the topic of active research. For
aforementioned reasons, we mainly focus on 2-point statistics, leaving the
possibility to include higher order CFs in the future work. The simplest 2-point
probability function ($S_2$ or autocorrelation) actually arise from small angle
scattering experiments and measures the probability that both ends of line
segment with a given length fall into the same phase. Other types include the
probability of a whole line segment to fall into the phase ($L_2$ function) or
to lie on the interface between phases ($F_{ss}$ function), and mainly originated
from aforementioned bounds on physical properties such as permeability and
elasticity. The general idea of computation of different 2-point CFs is shown in
\cref{fig:functions}. All major details and numerous analytical cases can be
found in seminal work of Torquato \cite{Torquato_book}. To all CFs described in this
book and applicable to digital images we shall refer to as classical correlation
functions.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{images/corr.png}
  \caption[]{A schematic depiction of a binary porous media (pores are shown in
    color) with examples of positive events for surface $F_{ss}$, $F_{sv}$ and
    two-point probability $S_2$ correlation functions. The zoomed in area
    represents the difference between the true ``continuous'' interface in
    between pore and solid phases with pixelized ``digital'' interface emerging
    due to limited resolution of digital images.}
  \label{fig:functions}
\end{figure}

@@

\code{CorrelationFunctions.jl} is capable to compute all aforementioned
correlation functions. Where possible we provide two implementations for each
CF. The first implementation computes a correlation function in specific
predefined directions (i.e., orientation of test vectors thrown into a sample is
fixed) \cite{jiao2014chawla,EPL1}.  There are one predefined direction in 1D
case, four directions in 2D case (two orthogonal and two diagonal) and thirteen
directions in 3D case (three orthogonal, six diagonal within each orthogonal
plane and four diagonal). The line segment scanning approach is slower for some
CFs but requires much less memory resources. The full map approach establishes
correlations between all points on the image is usually faster.  Both
implementations are contained in \code{Directional} and \code{Map} modules of
the package, correspondingly.

With the help of the package one can compute CFs in periodic and non-periodic
modes.  Periodic mode assumes that an input sample is periodically continued to
the whole Euclidean space. Non-periodic mode assumes zero-padded input.

@@

For surface-surface and surface-void functions we use an edge detection filter
\cite{samarin2023robust} and then calculate either autocorrelation of the edge
or cross-correlation of the edge and the void phase. The edge is extracted by
convolving an input with a short high-pass filter $H$. It has a width $7$ and
all its coefficients with exception of the central coefficient are inversely
proportional to distance from the center:
\begin{equation}
  \begin{aligned}
    H_{ij} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2}} & \quad i \ne 3, j \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right. \\
    H_{ijk} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2 + (k-3)^3}} & \quad i \ne 3, j \ne 3, k
      \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right.
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $C$ is such that all coefficients of $H'$ sum to zero, and $S$ equals to
$30.45849$ in 2D and $172.96232$ in 3D case. The parameters of the filter were adjusted
based on comparison against exact computations of surface functions on smooth-boundary
sets \cite{Postnicov2023}.

@@

The interface between phases must not change very rapidly for the edge filter to
work correctly. Therefore, the algorithms for surface functions require input
images to be in good resolution. A criterion for goodness can be expressed with
$C_{\alpha}$ parameter. Let $\alpha \in [0, 1]$ and $f(x)$ be your
(band-limited) input image ($x \in \mathbb{R}^2$ or $\mathbb{R}^3$). We define
$C_\alpha$ as follows:
\begin{align*}
  f_0(x) &= f(x) - \langle f(x) \rangle \\
  C_\alpha &= \frac{\int_{-\alpha\omega}^{\alpha\omega} |\hat{f_0}(z)|^2
    dz}{\int_{-\omega}^{\omega} |\hat{f_0}(z)|^2 dz}
\end{align*}
where $\hat{f_0}$ is a Fourier transform of $f_0$ and $\omega$ is a folding
frequency which depends only on resolution of the image. Our criterion is that
if $C_{0.5} > 0.97$ then the image is suitable for our algorithms which compute
$F_{ss}$ and $F_{sv}$.

Our library does not have an implementation for lineal-path maps, although we
know that such implementations exist \cite{havelkafast}. The reason is a very
slow computation even on GPUs, plus the resulting map is not for all points on
the image, but within a spherical volume (due to rotation of the 3D image during
lineal-path evaluation). Moreover, linear scan approach is orders of magnitude
faster for $L_2$.

All described algorithms contain parallelizable operations such as fast Fourier
transform, connected components labeling and image filtering, hence they are
perfectly suited for execution on GPUs and provide very fast computations (see
wall times below).

\subsection{Directional scanning with line segment}
\label{sec:scan}
All algorithms described above (full correlation maps) are executed on the whole
input and hence require a lot of RAM memory. The scan approach deals with memory
limitations at the cost of higher execution time (except for $L_2$). With
line-segment scanning we select a list of predefined directions and cut all
one-dimensional slices from the input along those directions. Then for each
direction we use an appropriate algorithm in \cref{sec:map} to calculate a
correlation function just for that single slice. When all slices are processed
we calculate the resulting directional or averaged CF using the formula:
\begin{equation*}
  F^{\bm{d}}(r) = \frac{\sum\limits_i F^{\bm{d}}_i(r) N_i(r)}{\sum\limits_i N_i(r)}
\end{equation*}
Here $F^{\bm{d}}(r)$ is a correlation function computed with respect to a
direction $\bm{d}$, $F^{\bm{d}}_i(r)$ is a ``partial'' correlation
function calculated for a $i$-th slice and $N_i(r)$ is a total number of trials
for a correlation length $r$ and slice $i$. Some algorithms require
preprocessing stages before cutting the input in slices, e.g., image filtering
for edge detection or connected components labeling performed on the whole
array, not on slices.

An algorithm for computation of linear-path function for one-dimensional slice
is presented below. Here $a$ is a one-dimensional input array, $phase$ can be
void, solid or some other phase, and $periodic$ is a boolean value which equals
to true if we compute $L_2$ in periodic mode and false otherwise. The efficiency
of the proposed algorithm is $O(n)$ where $n$ is the length of the array.
\begin{algorithmic}[1]
  \Procedure{l2}{$A, phase, periodic$}
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \State $runs \gets countruns(A, phase)$
    \ForAll{$run \in runs$}
      \State $updateruns(result, run)$
    \EndFor
    \If{$periodic$}
      \State $first \gets first(A)$
      \State $last \gets last(A)$
      \If{$first = last$}
        \State $updateperiodic(result, first, last)$
      \EndIf
    \EndIf
    \State \textbf{return} $success$ divided by number of trials
    (see \cref{sec:number-of-trials})
  \EndProcedure
  \\
  \Procedure{countruns}{$A, phase$}
    \State $runslist \gets [\ ]$
    \State $runs \gets 0$
    \ForAll{$x \in A$}
      \If{$x = phase$}
        \State $runs \gets runs + 1$
      \ElsIf{$runs \ne 0$}
        \Comment Add $runs$ to the accumulator
        \State $runslist \gets runs:runslist$
        \State $runs \gets 0$
      \EndIf
    \EndFor
    \If{$runs \ne 0$}
      \State $runslist \gets runs:runslist$
    \EndIf
    \State \textbf{return} $runslist$
  \EndProcedure
  \\
  \Procedure{updateruns}{$A, runs$}
    \State $r \gets runs$
    \For{$idx \gets 0, runs - 1$}
      \State $A[idx] \gets A[idx] + r$
      \State $r \gets r - 1$
    \EndFor
  \EndProcedure
  \\
  \Procedure{updateperiodic}{$A, first, last$}
    \State $sum \gets first + last$
    \For{$idx \gets 0, sum - 1$}
      \State $up \gets \min(idx, first, last, sum - idx)$
      \State $A[idx] \gets A[idx] + up$
    \EndFor
  \EndProcedure
\end{algorithmic}

Cluster function can be calculated using the same technique as in map approach,
but if the number of clusters is high, a naïve $O(n^2)$ algorithm may be more
preferred than $O(MC \log M)$ algorithm. Such an algorithm is present below:
\begin{algorithmic}[1]
  \Procedure{c2}{$A, phase, periodic$}
    \State $A' \gets I^{(phase)}(A)$
    \State $CC \gets label\_components(A', periodic)$
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \ForAll{$slice \in slices(CC)$}
      \For{$i = 0,len-1$}
        \State $end \gets len-1$
        \If{$periodic$}
          \State $end \gets end + i$
        \EndIf
        \For{$j = i, end$}
          \If{$slice[i] = slice[j \mod len] \ne 0$}
            \Comment $0$ is a label for background (phase not equal to $phase$).
            \State $dst \gets |(j \mod len) - i|$
            \State $result[dst] \gets result[dst] + 1$
          \EndIf
        \EndFor
      \EndFor
    \EndFor
    \State \textbf{return} $result$ divided by number of trials
  \EndProcedure
\end{algorithmic}

Before computation of chord length function an input array must be pre-processed
to find an interface for the phase of interest. For this purpose we use
distance transform function defined as follows:
\begin{equation}
  \mathcal{D}(\bm{x})= \left\{
  \begin{array}{ll}
    0 & \quad \bm{x} \in \text{solid phase} \\
    \min\limits_{y \in \text{solid phase}} \rho(\bm{x},\bm{y}) & \quad \text{otherwise}
  \end{array}
\right. \label{eq:distance-transform}
\end{equation}
There are many algorithms for distance transform, including algorithms performed
in linear time \cite{DT}.In \code{chord\_length} function we firstly apply
distance transform to an input to find the interface and then accumulate chord
length in per-slice manner. Finally, chord lengths can be binned in a histogram.
\begin{algorithmic}[1]
  \Procedure{chord\_length}{$A, phase$}
    \State $A' \gets I^{(phase)}(A)$
    \State $A_E \gets \mathcal{D}(A') = 1$
    \Comment We consider elements for which the distance transform equals to $1$
    to be part of the interface
    \State $chord\_list \gets [\ ]$
    \ForAll{$(S', S_E) \in zip(slices(A'), slices(A_E))$}
      \Comment iterate through all one-dimensional slices in $A'$ and
      corresponding slices in $A_E$
      \State $l \gets 0$
      \State $maybe\_chord \gets False$
      \For{$idx \gets 0, length(S')-1$}
        \If{$S_E[idx]$}
          \Comment Interface found
          \If{$l > 1\ and\ maybe\_chord$}
            \Comment Chord found
            \State $chord\_list \gets l:chord\_list$
          \EndIf
          \State $l \gets 0$
          \State $maybe\_chord \gets True$
        \ElsIf{$\neg S'[idx]$}
          \Comment We are not in the phase of interest
          \State $maybe\_chord \gets False$
        \EndIf
      \EndFor
    \EndFor
    \State \textbf{return} $chord\_list$ or $histogram(chord\_list)$
  \EndProcedure
\end{algorithmic}
This implementation also works in linear time.

@@

\subsection{Comparison against analytical solutions}
To test out package we generate a realization of random overlapping disks
(\cref{fig:overlapping-disks}) with dimensions $10000 \times 10000$ and
parameters $R = 40\ px$ and $\lambda = 10^{-4}\ px^{-2}$ and compare calculated
correlation functions with their theoretical values.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{images/disks.png}
  \includegraphics[width=0.45\linewidth]{images/balls.png}
  \caption[]{Images of overlapping disks and balls used in testing of our library.}
  \label{fig:overlapping-disks}
\end{figure}

For a 3D test, we generate a realization of random overlapping spheres
(\cref{fig:overlapping-disks}) with dimensions $500 \times 500 \times 500$ and
parameters $R = 10\ px$ and $\lambda = 10^{-4}\ px^{-3}$ and compare calculated
correlation functions with their theoretical values.

The results of comparison of computational results against analytical solutions 
are present on \cref{fig:verification}. It is clear from the test shown that CFs 
are computed with high accuracy; minor deviations are due to finite size of our
Poisson realizations (i.e., larger images reduce deviations \cite{samarin2023robust}).

@@

\section{Applications and benchmarks}
After the verification of our computational approach based on analytical
solution, in this section we apply \code{CorrelationFUnctions.jl} to compute CFs
for a variety of real 2D (SEM) and 3D (XCT) images. We also report benchmark on
execution time and compare GPU and CPU implementations.

\subsection{Examples of application to experimental images}
\label{sec:examples}
A summary which contains information about type of the images, dimensions and resolution
can be found in \cref{tab:samples}. Correlation functions of XCT images can be
seen on \cref{fig:real-xct} and correlation functions of SEM images are on
\cref{fig:real-sem}.

\begin{table}[!pt]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Sample name & Image type & Dimensions (pixels) & Image resolution ($\mu m$) & $C_{0.5}$\\
    \hline
    Ceramic 1   & XCT & $500 \times 500 \times 500$ & 2.25 & 0.9823 \\
    Ceramic 2   & XCT & $500 \times 500 \times 500$ & 2.25 & 0.9890 \\
    Ceramic 3   & XCT & $500 \times 500 \times 500$ & 2.25 & 0.9712 \\
    Sandstone 1 & SEM &  $1280 \times 869$ & 0.10 & 0.9703 \\
    Sandstone 2 & SEM &  $1280 \times 869$ & 0.05 & 0.9897 \\
    Carbonate 1 & SEM &  $1024 \times 691$ & 0.06 & 0.9721 \\
    \hline
  \end{tabular}
  \caption{Summary of samples used for calculation of correlation functions.}
  \label{tab:samples}
\end{table}

@@

\subsection{Execution times for our algorithms}
\label{sec:efficiency}
To measure execution times of our algorithms we use a machine with the following
configuration:
\begin{itemize}
\item \textbf{CPU}: AMD Ryzen 7 5800X
\item \textbf{Memory}: 32~Gb DDR4 memory
\item \textbf{GPU}: Nvidia Tesla V100 with 32~Gb video memory
\end{itemize}

We generate two (resp. three) dimensional square (resp. cubic) binary arrays
with different sides. An array used for measurements contains two solid disks
(resp. balls) surrounded by void phase.

To measure efficiency of our code we generate six arrays with a side of a square
$s = 1000, 2000, \dots, 6000$ pixels for two-dimensional case and six arrays
with a side of a cube $s = 50, 100, \dots, 300$ voxels for three-dimensional
case and calculate execution time for each function. All functions are
calculated using periodic boundary conditions. Functions which use per-slice
directional approach are calculated along axial directions.

Functions which calculate full correlation maps are suitable for running on CPU
and GPU, so we provide execution times for both. The cluster function still uses
an algorithm for closest component labeling which works only on CPU, so there is
a potentional for improvement for this function.

The measurements of execution time is on \cref{fig:timings}. As you can see, GPU
versions (when available) are faster than CPU versions and limited only by
amount of video memory. Functions in module \code{Directional} (which utilize
per-slice directional approach) are the slowest among other implementations, but
require less memory.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{images/timing-image.png}
  \caption[]{One of 2D image used in timing measurements.}
  \label{fig:valuenoise}
\end{figure}

>>>>

\bibliographystyle{elsarticle-num}
\journal{Computational Material Science}

@@

\author[Smith]{Vasily~Postnicov}
\author[Smith,MIT]{Aleksei~Samarin}
\author[Smith]{Marina ~V. Karsanina}
\author[Aus]{Martieu ~Gravey}
\author[Smith]{Aleksey~Khlyupin}
\author[Smith]{Kirill ~M. Gerke}
\ead{kg@ifz.ru}

\address[Smith]{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 107031, Russia}
\address[MIT]{Computational Mathematics and Cybernetics, Lomonosov Moscow State
  University, Moscow, 119991, Russia}
\address[Aus]{Institute for Interdisciplinary Mountain Research, Austrian Academy of Sciences, Innsbruck, 6020, Austria}

@@

There are different types of correlation functions that in general describe some
probabilities. The main parameter of any correlation function is the number of
points that are utilized to evaluate such a probability. While so-called
$n$-point probability function \cite{Torquato_book} will completely describe any
structure in the limit of $n \rightarrow$ number of voxels/pixels on the image,
computing or storing such a function is not practical. Based on information
content, it was shown that increasing the number of point $n > 2$ only
marginally improves the quality of the structure quantification and these
improvements decay with increasing $n$ \cite{yao1993high,Gommes2}. For stochastic reconstruction purposes
computation of higher-order statistics will, arguably, not be balanced by
increased number of points, but this is the topic of active research \cite{skolnick2021,cherkasovIC}. For
aforementioned reasons, we mainly focus on 2-point statistics, leaving the
possibility to include higher order CFs in the future work. The simplest 2-point
probability function ($S_2$ or autocorrelation) actually arise from small angle
scattering experiments and measures the probability that both ends of line
segment with a given length fall into the same phase. Other types include the
probability of a whole line segment to fall into the phase ($L_2$ function) or
to lie on the interface between phases ($F_{ss}$ function), and mainly originated
from aforementioned bounds on physical properties such as permeability and
elasticity. The general idea of computation of different 2-point CFs is shown in
\cref{fig:functions}. All major details and numerous analytical cases can be
found in seminal work of Torquato \cite{Torquato_book}. To all CFs described in this
book and applicable to digital images we shall refer to as classical correlation
functions.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{images/figa1.jpg}
  \caption[]{A schematic depiction of a binary porous media (pores are shown in
    color) with examples of positive events for surface $F_{ss}$, $F_{sv}$ and
    two-point probability $S_2$ correlation functions. The zoomed in area
    represents the difference between the true ``continuous'' interface in
    between pore and solid phases with pixelized ``digital'' interface emerging
    due to limited resolution of digital images.}
  \label{fig:functions}
\end{figure}

@@

\code{CorrelationFunctions.jl} is capable to compute all aforementioned
correlation functions. Where possible we provide two implementations for each
CF. The first implementation computes a correlation function in specific
predefined directions (i.e., orientation of test vectors thrown into a sample is
fixed) \cite{jiao2014chawla,EPL1}.  There are one predefined direction in 1D
case, four directions in 2D case (two orthogonal and two diagonal) and thirteen
directions in 3D case (three orthogonal, six diagonal within each orthogonal
plane and four diagonal). All directions available are schematically represented in \cref{fig:directions}. The line segment scanning approach is slower for some
CFs but requires much less memory resources. The full map approach establishes
correlations between all points on the image is usually faster.  Both
implementations are contained in \code{Directional} and \code{Map} modules of
the package, correspondingly.

With the help of the package one can compute CFs in periodic and non-periodic
modes.  Periodic mode assumes that an input sample is periodically continued to
the whole Euclidean space. Non-periodic mode treats the input as zero-padded image.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{images/figa_direction.jpg}
  \caption[]{Available directions to compute directional correlation functions in 2D and 3D cases with line scanning approach.}
  \label{fig:directions}
\end{figure}

@@

For surface-surface and surface-void functions we use an edge detection filter
\cite{samarin2023robust} and then calculate either autocorrelation of the edge
or cross-correlation of the edge and the void phase. The edge is extracted by
convolving an input with a short high-pass filter $H$. It has a width $7$ and
all its coefficients with exception of the central coefficient are inversely
proportional to distance from the center:
\begin{equation}
  \begin{aligned}
    H_{ij} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2}} & \quad i \ne 3, j \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right. \\
    H_{ijk} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2 + (k-3)^3}} & \quad i \ne 3, j \ne 3, k
      \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right.
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $C$ is such that all coefficients of $H'$ sum to zero, and $S$ equals to
$30.45849$ in 2D and $172.96232$ in 3D case. The parameters of the filter were adjusted
based on comparison against exact computations of surface functions on smooth-boundary
sets \cite{postcov23}.

@@

The interface between phases must not change rapidly for the edge filter to
work correctly. Therefore, the algorithms for surface functions require input
images to be with an apropriate resolution. A criterion for goodness can be expressed with
$C_{\alpha}$ parameter. Let $\alpha \in [0, 1]$ and $f(x)$ be
(band-limited) input image ($x \in \mathbb{R}^2$ or $\mathbb{R}^3$). We define
$C_\alpha$ as follows:
\begin{align*}
  f_0(x) &= f(x) - \langle f(x) \rangle \\
  C_\alpha &= \frac{\int_{-\alpha\omega}^{\alpha\omega} |\hat{f_0}(z)|^2
    dz}{\int_{-\omega}^{\omega} |\hat{f_0}(z)|^2 dz}
\end{align*}
where $\hat{f_0}$ is a Fourier transform of $f_0$ and $\omega$ is a folding
frequency which depends only on resolution of the image. We have previously shown that criterion 
$C_{0.5} > 0.97$ holds then the image is suitable for robust $F_{ss}$ and $F_{sv}$ evaluation \cite{samarin2023robust}.
Note that function \code{lowfreq\_energy\_ratio} that computes $C_{0.5}$ criterion for any input binary image is the part of \code{CorrelationFunctions.jl} package.

Our library does not have an implementation for lineal-path and chord-length maps, although we
know that such implementations exist \cite{turner2016efficient,Havelka}. The reason is a very
slow computation even on GPUs, plus the resulting map is not for all points on
the image, but within a spherical volume (due to rotation of the 3D image during
lineal-path evaluation). Moreover, linear scan approach is orders of magnitude
faster for $L_2$.

Majority of described algorithms contain parallelizable operations such as fast Fourier
transform, connected components labeling and image filtering, hence they are
perfectly suited for execution on GPUs and provide very fast computations (see
wall times below).

\subsection{Directional scanning with line segment}
\label{sec:scan}
All algorithms described above (full correlation maps) are executed on the whole
input image and hence require a lot of RAM memory. The scan approach deals with memory
limitations at the cost of higher execution time (except for $L_2$). With
line-segment scanning we select a list of predefined directions and cut all
one-dimensional slices from the input along those directions. Then for each
direction, as shown in  \cref{fig:directions}, we use an appropriate algorithm in \cref{sec:map} to calculate a
correlation function just for that single slice. For this reason we do not repeat algorithms for $S_2$, $\rho_{ij}$, $F_{ss}$ and $F_{sv}$.
When all slices are processed we calculate the resulting directional or averaged CF using the formula:
\begin{equation*}
  F^{\bm{d}}(r) = \frac{\sum\limits_i F^{\bm{d}}_i(r) N_i(r)}{\sum\limits_i N_i(r)}
\end{equation*}
Here $F^{\bm{d}}(r)$ is a correlation function computed with respect to a
direction $\bm{d}$, $F^{\bm{d}}_i(r)$ is a ``partial'' correlation
function calculated for a $i$-th slice and $N_i(r)$ is a total number of trials
for a correlation length $r$ and slice $i$. Some algorithms require
preprocessing stages before cutting the input in slices, e.g., image filtering
for edge detection or connected components labeling performed on the whole
array, not on slices.

An algorithm for computation of linear-path function for one-dimensional slice
is presented below. Here $A$ is a one-dimensional input array, $phase$ can be
void, solid or some other phase, and $periodic$ is a boolean value which equals
to true if we compute $L_2$ in periodic mode and false otherwise. The efficiency
of the proposed algorithm is $O(n)$ where $n$ is the length of the array.
\begin{algorithmic}[1]
  \Procedure{l2}{$A, phase, periodic$}
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \State $runs \gets countruns(A, phase)$
    \ForAll{$run \in runs$}
      \State $updateruns(result, run)$
    \EndFor
    \If{$periodic$}
      \State $first \gets first(A)$
      \State $last \gets last(A)$
      \If{$first = last$}
        \State $updateperiodic(result, first, last)$
      \EndIf
    \EndIf
    \State \textbf{return} $success$ divided by number of trials
    (see \cref{sec:number-of-trials})
  \EndProcedure
  \\
  \Procedure{countruns}{$A, phase$}
    \State $runslist \gets [\ ]$
    \State $runs \gets 0$
    \ForAll{$x \in A$}
      \If{$x = phase$}
        \State $runs \gets runs + 1$
      \ElsIf{$runs \ne 0$}
        \Comment Add $runs$ to the accumulator
        \State $runslist \gets runs:runslist$
        \State $runs \gets 0$
      \EndIf
    \EndFor
    \If{$runs \ne 0$}
      \State $runslist \gets runs:runslist$
    \EndIf
    \State \textbf{return} $runslist$
  \EndProcedure
  \\
  \Procedure{updateruns}{$A, runs$}
    \State $r \gets runs$
    \For{$idx \gets 0, runs - 1$}
      \State $A[idx] \gets A[idx] + r$
      \State $r \gets r - 1$
    \EndFor
  \EndProcedure
  \\
  \Procedure{updateperiodic}{$A, first, last$}
    \State $sum \gets first + last$
    \For{$idx \gets 0, sum - 1$}
      \State $up \gets \min(idx, first, last, sum - idx)$
      \State $A[idx] \gets A[idx] + up$
    \EndFor
  \EndProcedure
\end{algorithmic}

Cluster function can be calculated using the same technique as in map approach,
but if the number of clusters is high, a naïve $O(n^2)$ algorithm may be
preferred to $O(MC \log M)$ full map algorithm. Such an approach is present below:
\begin{algorithmic}[1]
  \Procedure{c2}{$A, phase, periodic$}
    \State $A' \gets I^{(phase)}(A)$
    \State $CC \gets label\_components(A', periodic)$
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \ForAll{$slice \in slices(CC)$}
      \For{$i = 0,len-1$}
        \State $end \gets len-1$
        \If{$periodic$}
          \State $end \gets end + i$
        \EndIf
        \For{$j = i, end$}
          \If{$slice[i] = slice[j \mod len] \ne 0$}
            \Comment $0$ is a label for background (phase not equal to $phase$).
            \State $dst \gets |(j \mod len) - i|$
            \State $result[dst] \gets result[dst] + 1$
          \EndIf
        \EndFor
      \EndFor
    \EndFor
    \State \textbf{return} $result$ divided by number of trials
  \EndProcedure
\end{algorithmic}

Before computation of chord length function an input array must be pre-processed
to find an interface for the phase of interest. For this purpose we use
distance transform function defined as follows:
\begin{equation}
  \mathcal{D}(\bm{x})= \left\{
  \begin{array}{ll}
    0 & \quad \bm{x} \in \text{solid phase} \\
    \min\limits_{y \in \text{solid phase}} \rho(\bm{x},\bm{y}) & \quad \text{otherwise}
  \end{array}
\right. \label{eq:distance-transform}
\end{equation}
There are many algorithms for distance transform, including algorithms performed
in linear time \cite{DT}.In \code{chord\_length} function we firstly apply
distance transform to an input to find the interface and then accumulate chord
length in per-slice manner. Finally, chord lengths can be binned to a histogram.
\begin{algorithmic}[1]
  \Procedure{chord\_length}{$A, phase$}
    \State $A' \gets I^{(phase)}(A)$
    \State $A_E \gets \mathcal{D}(A') = 1$
    \Comment We consider elements for which the distance transform equals to $1$
    to be part of the interface
    \State $chord\_list \gets [\ ]$
    \ForAll{$(S', S_E) \in zip(slices(A'), slices(A_E))$}
      \Comment iterate through all one-dimensional slices in $A'$ and
      corresponding slices in $A_E$
      \State $l \gets 0$
      \State $maybe\_chord \gets False$
      \For{$idx \gets 0, length(S')-1$}
        \If{$S_E[idx]$}
          \Comment Interface found
          \If{$l > 1\ and\ maybe\_chord$}
            \Comment Chord found
            \State $chord\_list \gets l:chord\_list$
          \EndIf
          \State $l \gets 0$
          \State $maybe\_chord \gets True$
        \ElsIf{$\neg S'[idx]$}
          \Comment We are not in the phase of interest
          \State $maybe\_chord \gets False$
        \EndIf
      \EndFor
    \EndFor
    \State \textbf{return} $chord\_list$ or $histogram(chord\_list)$
  \EndProcedure
\end{algorithmic}
This implementation also works in linear time.

@@

\subsection{Comparison against analytical solutions}
To test out package we generate a realization of random overlapping disks
(\cref{fig:overlapping-disks}) with dimensions $10000 \times 10000$ and
parameters of radii $R = 40\ pixels$ and $\lambda = 10^{-4}\ pixels^{-2}$ and then compare calculated
correlation functions with their theoretical values.
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{images/disks.png}
  \includegraphics[width=0.45\linewidth]{images/balls.png}
  \caption[]{Images of overlapping disks and balls used in testing of our library.}
  \label{fig:overlapping-disks}
\end{figure}

For a 3D test, we generate a realization of random overlapping spheres
(\cref{fig:overlapping-disks}) with dimensions $500 \times 500 \times 500$ and
$R = 10\ voxels$ and $\lambda = 10^{-4}\ voxels^{-3}$ and then compare calculated
correlation functions with analytical solutions.

The results of comparison of computational results against analytical solutions 
are present on \cref{fig:verification}. It is clear from the test shown that CFs 
are computed with high accuracy; minor deviations are due to finite size of our
Poisson realizations (i.e., larger images reduce deviations \cite{samarin2023robust}).
The package contains automatic tests based on these comparisons.

@@

\section{Applications and benchmarks}
After the verification of our computational approach based on analytical
solution, in this section we apply \code{CorrelationFUnctions.jl} to compute CFs
for a variety of real 2D (SEM) and 3D (XCT) images. We also report benchmarks on
execution time and compare the efficiency of GPU and CPU implementations.

\subsection{Examples of application to experimental images}
\label{sec:examples}
For testing purposes we have gathered an assortment of different porous materials including XCT images of
artificial ceramics \cite{ceramics} and natural soil composite \cite{Geoderma2018}, segmented into numerous phases,
and SEM images of oil-bearing rocks such as sandstones and carbonate \cite{samarin2023robust}.
A summary of this library of 2D and 3D images, including dimensions and resolutions,
can be found in \cref{tab:samples}.

Correlation functions of XCT images can be seen on \cref{fig:real-xct} and correlation functions of SEM images are on
\cref{fig:real-sem}. Cross-correlation functions for a multiphase XCT image is on \cref{fig:real-multiphase}.

\begin{table}[!pt]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Sample name & Image type & Dimensions (pixels) & Image resolution ($\mu m$) & $C_{0.5}$\\
    \hline
    Ceramic 1   & XCT & $500 \times 500 \times 500$ & 2.25 & 0.9823 \\
    Ceramic 2   & XCT & $500 \times 500 \times 500$ & 2.25 & 0.9890 \\
    Ceramic 3   & XCT & $500 \times 500 \times 500$ & 2.25 & 0.9712 \\
    Sandstone 1 & SEM &  $1280 \times 869$ & 0.10 & 0.9703 \\
    Sandstone 2 & SEM &  $1280 \times 869$ & 0.05 & 0.9897 \\
    Carbonate 1 & SEM &  $1024 \times 691$ & 0.06 & 0.9721 \\
    \hline
    Soil (multiphase sample) & XCT & $304 \times 304 \times 304$ &
    \textcolor{red}{2.25} & \\
    \hline
  \end{tabular}
  \caption{Summary of samples used for calculation of correlation functions.}
  \label{tab:samples}
\end{table}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{samples/multiphase.png}
  \includegraphics[width=0.45\linewidth]{samples/multiphase-cc.png}
  \caption[]{Cross-correlation functions for multiphase sample (soil) which has
    4 distinct phases, labeled as 1-4.}
  \label{fig:real-multiphase}
\end{figure}

\subsection{Execution times for our algorithms}
\label{sec:efficiency}
To measure execution times for developed computational algorithms we use a machine with the following
configuration:
\begin{itemize}
\item \textbf{CPU}: AMD Ryzen 7 5800X
\item \textbf{Memory}: 32~GB DDR4 memory
\item \textbf{GPU}: Nvidia Tesla V100 with 32~GB video memory
\end{itemize}

We generate 2 and 3-dimensional square/cubical binary arrays
with different width (see insets on \cref{fig:timings}). An array used for measurements contains two solid disks
or balls surrounded by void phase. This configuration is a compromise between ease of
image scaling and complexity of the structure (e.g., full map $C_2$ computational time
is not compromised by too many clusters).

We generated six arrays with dimensions of
$s = 1000, 2000, \dots, 6000$ pixels for two-dimensional case and six arrays
with a side of a cube $s = 50, 100, \dots, 300$ voxels for three-dimensional
case and calculated execution time for each correlation function. All functions are
calculated using periodic boundary conditions. Functions which use per-slice
directional approach were calculated along major axial directions only.

Functions which calculate full correlation maps are suitable for running on both CPU
and GPU, so we provide execution times for both. The cluster function uses
an algorithm for closest component labeling which has CPU-only implementation (thus,
leaving a room for future improvements). Directional computations are performed exclusively
on CPUs.

All measurements of execution time are presented on \cref{fig:timings}. It is immediately apparent that GPU
versions (where available) are much faster than CPU versions and limited only by
amount of video memory. Functions in module \code{Directional} (which utilize
per-slice directional approach) are the slowest among other implementations, but
require less memory. Depending on CF, our video card with 24 GBs of memory could handle a cube of
~$300^3$ voxels, whilst the workstation with 512-1024 GBs of RAM can handle virtually any XCT-obtained
image volume with any correlation length.
